Measurement Bias from Address Aliasing

### 2014-05-16 paper.tex
=Updates based on the feedback from Jørn.

    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchtectures. 
    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.

    We analyze a small program that was presented as a curiosity in previous work on \emph{measurement bias};
    Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.

@@

How fast does an algorithm run?
Is program A faster than program B on \emph{my} machine?
These types of questions are, perhaps surprisingly, not always trivial to answer.
The usual approach to quantify performance is to simply measure some metric, such as the wall time or number of cycles executed.
However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.

In other fields of science, accounting for sampling bias is well understood.
Trying to determine the political opinion of a country just by sampling a narrow geographical area will probably not be very accurate.
The same concept applies for software; measuring a metric once is analogous to sampling only a tiny geographical area in the set of possible system configurations.

There are many external variables that has the potential to affect system performance.
In particular, anything which can offset memory addresses of code or data at run-time can be significant.
Memory accesses interacts with low level hardware buffers, which can heavily favor certain memory layouts over others.
Consequentially, any environmental factor that can alter memory layout has a potential performance impact.
Failing to account for these effects lead to \emph{measurement bias}.

@@

Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowlege for potential optimization. % Could have a sentence with cite of Blind Optimization.
Usually, the root explanation will be that different memory layouts interact with various hardware buffers.
Certain mechanisms such as cache contention or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
Analyzing a program that exhibits strong bias from environment size, collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
Aliasing is an artifact of a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline.
In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous writes, comparing only the last 12 virtual address bits.
False dependencies incur performance penalties, and we show how this can explain many common cases of measurement bias.

@@

Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12 bit address suffix of 0x020 is the same in both. 
Dispite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
\end{description}

@@

% Generic intro to memory layout (with focus on stack/env interaction)
Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory. 
The stack is normally placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64 bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables.
There is still some level of stack alignment happening, independent of environment offsets. 
Normally, the stack is aligned to 16 bytes, which will be enforced by the compiler. % [ref]
Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data.

Note that there is no clear relationship between environment size and stack location with address randomization enabled.
However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurences of measurement bias indeed random.
% In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
% Moving to the lower end of the address space, we find the text, data and bss sections, which contains program code and statically allocated data.
% The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.

\begin{figure}[float=t]
  \caption{Organization of virtual memory for a 64 bit process.}
  \label{fig:virtualmemory}
  \begin{tikzpicture}[font=\footnotesize]

@@

\subsection{Analysis of micro-kernel}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by T. Mytkowicz et. al.~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}. 
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straight forward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.

@@

Notice the most extreme change from median to worst case is clearly the number of alias events.
If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
These counters are obviously not indicative of any causal relationship, and are ommitted.

% Need input here, does this make sense?
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270'000 to about 135'000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Intel:2013:Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports seem to have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
Our assumption is that this is a by-product of stalls generated from aliasing.

% More deep dive section (maybe separate headline...)
\subsection{The Offending Variables}
These numbers clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
For that we need to know the addresses of all variables at run time.
The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
Virtual memory layout of static data is decided at compile time, and we can find the location of these variables by looking at the symbol table in the ELF executable.
We find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044.\footnote{ELF symbol tables can be read using \texttt{readelf -s}}

Observing addresses of stack allocated data at run time is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
A small amount of assembly code was added to calculate the addresses of \texttt{g} and \texttt{inc}, outputing to stdout direcly using the \texttt{syscall} instruction.
Making no change to the stack allocation instructions and not affecting the addresses of static variables, the instrumented program have the exact same bias to environment size, free from observer effects.
We find that the spike in cycle count occurs precicely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe038 and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
The assembly output of gcc for the inner loop is shown below, indicating the conflicting load and store instructions.

@@

\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as address layout randomization. 
Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
Shown in Listing \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

@@

Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
In the previous sections we saw collisions between static data and stack, observing bias from external conditions that affected addresses of automatic variables.
Most dynamic memory is allocated on the \emph{heap}, which is managed by an \emph{allocator}.
A heap allocator is responsible for managing dynamic memory, and ultimately assigns the actual addresses of heap allocated variables at run time.
Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc, or some other library.
The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at run time.

\subsection{\texttt{malloc} will Alias by Default}
Acquiring dynamic memory at run time is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
\begin{itemize}
  \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
  Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.

  \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
  One drawback of this is that it requires the memory to be zero initialized, thus whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
\end{itemize}
The heap section starts at a relatively low address (close to the data section), growing upwards.
Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger -- for example 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}
This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.

@@

The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12 bits.
By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
To analyse performance for different addresses input and output buffers, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset the input arguments when calling \texttt{convolution}.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
This constant over head can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        convolution(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation is calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.

@@

\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
\begin{figure*}[t]
  \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}. Offset 0 means equal 12 bit address suffix, which is default behavior for \texttt{mmap} allocations. Input size $n=2^{20}$.}
  \label{fig:conv-default}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O2},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O3},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

@@

Listing \ref{lst:conv}, together with a small main method handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
Figure \ref{fig:conv-o2} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
This is an example of how false dependencies from address aliasing can cause a significant performance hit.
Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\%.
This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.

\subsection{Architectural Optimization}
It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.
You might also want to utilize the C99 keyword restrict to explicitly tell the compiler that the buffers are disjoint in memory.

@@

\subsection{Behavior of other Allocators}
As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
Heap allocation routines such as malloc and free are usually dynamically linked and resolved at run-time, which again represents a source of potential bias.
For the code examples presented earlier, the dynamic linker will resolve malloc to some version of glibc, which is the default on our system.
Variations between different versions of libc, or local configurations on parameters such as thresholds for when to use \texttt{mmap} over \texttt{brk}, could trigger aliasing cases.

In addition to glibc's ptmalloc, we look at the following alternatives.
\begin{itemize}
 \item Thread-Caching Malloc (ptmalloc) is developed by Google. It aims to be time and space efficient, and reduces lock contention for multithreaded applications. \cite{TCMalloc}
 \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments. \cite{JEMalloc}
 \item Hoard is another project that focuses extensively on improving performance for multithreaded programs. \cite{Berger:2000:Hoard}
\end{itemize}

@@

Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator. \footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
The addresses of two equally sized char buffers allocated with malloc are observed for size parameters 5120 and $2^{20}$ bytes. 
We do not care about performance, only the actual memory addresses returned and whether they alias or not.

For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
Interestingly, jemalloc and Hoard appears to never use brk, but always allocate to mmaped areas even for smaller sizes.
We see one example where one allocator yields aliasing buffers while another does not. 
Allocating $2 \times 5120 B$ returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.

Most malloc alternatives focuses heavily on performance in multithreaded environments.
Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
To our knowledge, there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
This dispite explicit recommendation from Intel that special purpose allocators should care about aliasing for optimization. [cite]

@@

\subsection{Dealing with Aliasing}
For the particular case we investigated, a good heuristic is to align heap segments “far apart” within the 4 KiB area of 12 bit suffixes.
More specifically, it appears that address suffixes of A and $\boldsymbol{y}$ are the most important to separate.

As discussed in section [x], aliasing cases like these can be accounted for in software using padding techniques.
A possible run time solution to adjust heap addresses can be realized as follows:

@@

Analyzing the example originally presented by Mytkowicz et al. \cite{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data resulted in aliasing for certain stack positions.
In this case, the external trigger of bias is variations in environment size, which in turn offset the virtual addresses of stack allocated variables.

>>>>

    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchitectures. 
    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.

    We analyze a small program that was presented as a curiosity in [others] previous work on \emph{measurement bias};
    Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.

@@

How fast does an algorithm run?
Is program A faster than program B on \emph{my} machine?
These types of questions are, perhaps surprisingly, not always trivial to answer.
The usual approach to quantify performance is to simply use some metric, such as the wall time or number of cycles executed.
However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.

In other fields of science, accounting for sampling bias is well understood.
Trying to determine the political opinion of a country just by sampling a narrow geographical area will probably not be very accurate.
The same concept applies for software; measuring once is analogous to sampling only a tiny geographical area in the set of possible system configurations.

There are many external variables that has the potential to affect system performance.
In particular, anything which can offset memory addresses of code or data at runtime can be significant.
Memory accesses interacts with low level hardware buffers, which can heavily favor certain memory layouts over others.
Consequentially, any environmental factor that can alter memory layout has a potential performance impact.
Failing to account for these effects lead to \emph{measurement bias}.

@@

Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowledge for potential optimization. % Could have a sentence with cite of Blind Optimization.
Usually, the explanation will be that different memory layouts interact with various hardware buffers.
Certain mechanisms such as cache contention or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
Analyzing a program that exhibits strong bias from environment size, collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
Aliasing is an artifact of a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline.
In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous writes, comparing only the last 12 virtual address bits.
False dependencies incur performance penalties, and we show how this can explain many common cases of measurement bias.

@@

Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
\end{description}

@@

% Generic intro to memory layout (with focus on stack/env interaction)
Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
The stack is normally placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables.
There is still some level of stack alignment happening, independent of environment offsets. 
Normally, the stack is aligned to 16 bytes, which will be enforced by the compiler. % [ref]
Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data.

Note that there is no clear relationship between environment size and stack location with ASLR enabled.
However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
% In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
% Moving to the lower end of the address space, we find the text, data and bss sections, which contains program code and statically allocated data.
% The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.

\begin{figure}[float=t]
  \caption{Organization of virtual memory for a 64-bit process.}
  \label{fig:virtualmemory}
  \begin{tikzpicture}[font=\footnotesize]

@@

\subsection{Microkernel Analysis}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by T. Mytkowicz et. al.~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}. 
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straight forward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.

@@

Notice the most extreme change from median to worst case is clearly the number of alias events.
If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
These counters are obviously not indicative of any causal relationship, and are omitted.

% Need input here, does this make sense?
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Intel:2013:Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports seem to have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
Our assumption is that this is a by-product of stalls generated from aliasing.

% More deep dive section (maybe separate headline...)
\subsection{The Offending Variables}
These numbers clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
For that we need to know the addresses of all variables at runtime.
The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
Virtual memory layout of static data is decided at compile time, and we can find the location of these variables by looking at the symbol table in the ELF executable.
We find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044.\footnote{ELF symbol tables can be read using \texttt{readelf -s}}

Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
A small amount of assembly code was added to calculate the addresses of \texttt{g} and \texttt{inc}, outputting to stdout directly using the \texttt{syscall} instruction.
Making no change to the stack allocation instructions and not affecting the addresses of static variables, the instrumented program have the exact same bias to environment size, free from observer effects.
We find that the spike in cycle count occurs precisely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe{\underline{038} and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
The assembly output of gcc for the inner loop is shown below, indicating the conflicting load and store instructions.

@@

\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR. 
Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
Shown in Listing \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

@@

Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
In the previous sections we saw collisions between static data and stack, observing bias from external conditions that affected addresses of automatic variables.
Most dynamic memory is allocated on the \emph{heap}, which is managed by an \emph{allocator}.
A heap allocator is responsible for managing dynamic memory, and ultimately assigns the actual addresses of heap allocated variables at runtime.
Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc, or some other library.
The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at runtime.

\subsection{\texttt{malloc} will Alias by Default}
Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
\begin{itemize}
  \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
  Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.

  \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
  One drawback of this is that it requires the memory to be zero initialized, thus whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
\end{itemize}
The heap section starts at a relatively low address (close to the data section), growing upwards.
Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger -- for example 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every mmap'ed address ends with 0x010}
This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.

@@

The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
To analyse performance for different addresses input and output buffers, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset the input arguments when calling \texttt{convolution}.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
This constant over head can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        convolution(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation is calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.

@@

\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
\begin{figure*}[t]
  \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Input size $n=2^{20}$.}
  \label{fig:conv-default}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O2},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O3},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

@@

Listing \ref{lst:conv}, together with a small main function handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
Figure \ref{fig:conv-o2} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
This is an example of how false dependencies from address aliasing can cause a significant performance hit.
Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\%.
This is a speedup on top of compiler optimization level O3, which is not able to account for aliasing effects.

\subsection{Architectural Optimization}
It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.
Another option is to utilize the C99 keyword \emph{restrict} to explicitly tell the compiler that the buffers are disjoint in memory.

@@

\subsection{Behavior of other Allocators}
As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
Heap allocation routines such as malloc and free are usually dynamically linked and resolved at runtime, which again represents a source of potential bias.
For the code examples presented earlier, the dynamic linker will resolve malloc to some version of glibc, which is the default on our system.
Variations between different versions of libc, or local configurations on parameters such as thresholds for when to use \texttt{mmap} over \texttt{brk}, could trigger aliasing cases.

In addition to glibc's ptmalloc, we look at the following alternatives.
\begin{itemize}
 \item Thread-Caching Malloc (tcmalloc) is developed by Google. It aims to be time and space efficient, and reduces lock contention for multithreaded applications. \cite{TCMalloc}
 \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments. \cite{JEMalloc}
 \item Hoard is another project that focuses extensively on improving performance for multithreaded programs. \cite{Berger:2000:Hoard}
\end{itemize}

@@

Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
The addresses of two equally sized char buffers allocated with malloc are observed for size parameters 5120 and $2^{20}$ bytes. 
We do not care about performance, only the actual memory addresses returned and whether they alias or not.

For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
Interestingly, jemalloc and Hoard appears to never use brk, but always allocate to memory mapped areas even for smaller sizes.
We see one example where one allocator yields aliasing buffers while another does not. 
Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.

Most malloc alternatives focuses heavily on performance in multithreaded environments.
Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
To our knowledge, there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
This despite explicit recommendation from Intel that special purpose allocators should care about aliasing for optimization. [cite]

@@

\subsection{Dealing with Aliasing}
For the particular case we investigated, a good heuristic is to align heap segments “far apart” within the 4 KiB area of 12-bit suffixes.
More specifically, it appears that address suffixes of A and $\boldsymbol{y}$ are the most important to separate.

As discussed in section [x], aliasing cases like these can be accounted for in software using padding techniques.
A possible runtime solution to adjust heap addresses can be realized as follows:

@@

By analyzing an example originally presented by Mytkowicz et al. \cite{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data resulted in aliasing for certain stack positions.
In this case, the external trigger of bias is variations in environment size, which in turn offset the virtual addresses of stack allocated variables.
